### 2025-02-15 causal-inference.qmd
20250214 - copyediting

One way to control for a variable—and what is typically considered the most rigorous control—when possible, is to randomize.
For example, you could randomly assign players to drink sports drink during a game versus water.
By randomly assigning players to each group, it is expected that, if the groups are large enough, the differences between the groups on all variables besides the [independent](#sec-experiment) and [dependent](#sec-experiment) variables will be approximately equal across both group; thus, any difference in the [dependent variable](#sec-experiment) is thought to be attributable to the manipulation of the [independent variable](#sec-experiment).
However, sometimes a variable is not able to be manipulated for practical or ethical reasons.
For instance, if you wanted to control for education level when examining the association between sports drink consumption and performance, you cannot randomly assign players to have lower educational levels (though you could possibly provide an intervention to increase the education levels of a subset of players).

@@

Matching means that you match people based on the control variable.
If you some people consume sports drink whereas others do not, matching would mean finding—for every player included in the study who consumes sports drink—an equivalent player in terms of education who does not consume sports drink.

>>>>

One way to control for a variable—and what is typically considered the most rigorous control—when possible, is to randomize (which would make it an [experiment](#sec-causalInferenceExperiment)).
For example, you could randomly assign players to consume sports drink during a game versus water.
By randomly assigning players to each group, it is expected that, if the groups are large enough, the differences between the groups on all variables besides the [independent](#sec-experiment) and [dependent](#sec-experiment) variables will be approximately equal across both groups; thus, any difference in the [dependent variable](#sec-experiment) is thought to be attributable to the manipulation of the [independent variable](#sec-experiment).
However, sometimes a variable is not able to be manipulated for practical or ethical reasons.
For instance, if you wanted to control for education level when examining the association between sports drink consumption and performance, you cannot randomly assign players to have lower educational levels (though you could possibly provide an intervention to increase the education levels of a subset of players).

@@

Matching means that you match people based on the control variable.
If some people consume sports drink whereas others do not, matching would mean finding—for every player included in the study who consumes sports drink—an equivalent player in terms of education who does not consume sports drink.
